---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.cap = "",
  out.width = "100%",
  fig.width = 8,
  fig.height = 4.2,
  dpi = 150
)
```

# bsfa

<!-- badges: start -->
[![R-CMD-check](https://github.com/franzmohr/bsfa/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/franzmohr/bsfa/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/franzmohr/bsfa/graph/badge.svg)](https://app.codecov.io/gh/franzmohr/bsfa)
[![License: GPL (>= 2)](https://img.shields.io/badge/license-GPL%20%28%3E%3D%202%29-blue.svg)](https://www.gnu.org/licenses/gpl-2.0)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

[![GitHub Sponsors](https://img.shields.io/badge/Sponsor-%E2%9D%A4-ea4aaa?logo=github-sponsors&logoColor=white)](https://github.com/sponsors/franzmohr)
[![Buy Me a Coffee](https://img.shields.io/badge/Buy%20Me%20a%20Coffee-FFDD00?logo=buymeacoffee&logoColor=black)](https://buymeacoffee.com/franzmohr)
<!-- badges: end -->

Bayesian estimation of stochastic frontier models in R.

The package implements Gibbs samplers with data augmentation for composed-error
frontier models in the tradition of van den Broeck, Koop, Osiewalski and Steel
(1994). Because the one-sided inefficiency term is sampled as a latent variable
rather than integrated out, efficiency scores come out of the sampler as draws.
`efficiency()` therefore reports genuine posterior distributions, where maximum
likelihood has to fall back on the Jondrow et al. (1982) conditional mean — a
point predictor of an unobserved quantity, with no comparable measure of
uncertainty attached.

At the time of writing there is no maintained general-purpose Bayesian
stochastic frontier package on CRAN. `sfaR`, `sfa`, `frontier`, `semsfa` and
`ssfa` are all maximum likelihood.

## Installation

```r
# install.packages("remotes")
remotes::install_github("franzmohr/bsfa")
```

## Usage

A model is built up step by step, in the manner of
[bvartools](https://github.com/franzmohr/bvartools): a constructor fixes the
data and the specification, and the `add_*` functions attach the remaining
blocks and then the posterior draws.

```{r usage}
library(bsfa)

set.seed(1234)
d <- sim_sf(n = 500, beta = c(1, 0.5, 0.3), sigma_v = 0.2, par_u = 4)

model <- create_sfmodel_exp(y ~ x1 + x2, data = d,
                            iterations = 5000, burnin = 2000)
model <- add_priors(model)
model <- add_initial_values(model)
model <- add_seed(model, 1234)
model <- add_posterior_coefficients(model)
model <- add_posterior_loglik(model)

summary(model)
```

The data were simulated with $\beta = (1, 0.5, 0.3)$, $\sigma_v = 0.2$ and
$\lambda = 4$. The slopes come back sharply; the intercept is the loose one,
because it has to be separated from the mean of the one-sided term using only
the asymmetry of the residuals.

The draws are attached to the same object rather than returned as a separate
result, and are stored as `coda` objects, one block per parameter. Printing the
model reports which blocks have been set, which is worth a glance before
committing to a long run.

```{r efficiency-plot}
plot(model, type = "efficiency")
```

The left panel is the summary usually reported. The right one is the reason to
look further: with one observation per unit there is very little information
about any individual score, and the bands cover much of the range the histogram
spreads over. That uncertainty is what the maximum likelihood route cannot
report.

```{r criteria}
selection_criteria(model)
```

## Model variants are classes

The distribution of the inefficiency term is carried by the model's class rather
than by an argument, so each variant gets its own methods:

| Constructor | Class | Inefficiency |
|---|---|---|
| `create_sfmodel_exp()` | `sfmodel_exp` | exponential, rate `lambda` |
| `create_sfmodel_hn()` | `sfmodel_hn` | half-normal, scale `sigma_u` |
| `create_sfmodel4_exp()` | `sfmodel4_exp` | four-component, exponential |
| `create_sfmodel4_hn()` | `sfmodel4_hn` | four-component, half-normal |

That split is not cosmetic. The prior on the inefficiency term is elicited from
a prior median efficiency `r_star`, and the two distributions map that anchor
differently, so the argument is named after the parameter it governs:

```r
model <- add_priors(model, lambda = list(r_star = 0.85))   # sfmodel_exp
model <- add_priors(model, sigma_u = list(r_star = 0.85))  # sfmodel_hn
```

For the exponential model the mapping is exact. With a gamma prior of shape 1
and rate `c = -log(r_star)` on the rate parameter, the implied marginal prior
density of `u` is `c / (u + c)^2`, whose median is `c`, so the prior median of
`exp(-u)` is exactly `r_star`. For the half-normal model the same target is
matched only in expectation.

## What is implemented

| | |
|---|---|
| Inefficiency | exponential, half-normal |
| Frontier | production, cost |
| Data | cross-section, time-invariant panel (Pitt and Lee, 1981), four-component panel (Kumbhakar, Lien and Hjalmarsson, 2014) |
| Output | posterior draws of coefficients, variances and unit-level efficiency |
| Inference | `summary()`, `plot()`, `efficiency()`, `selection_criteria()` (LL, AIC, BIC, HQ, WAIC) |

## The four-component model

`create_sfmodel4_exp()` and `create_sfmodel4_hn()` split the disturbance into a
unit effect, persistent inefficiency, noise and transient inefficiency. It needs
panel data, so `id` is required.

```{r four-component}
dp <- sim_sf4(n = 120, n_time = 8, beta = c(1, 0.5, 0.3))

m4 <- create_sfmodel4_exp(y ~ x1 + x2, data = dp, id = "id",
                          iterations = 5000, burnin = 2000)
m4 <- add_posterior_coefficients(add_priors(m4))

head(efficiency(m4, type = "persistent"), 3)   # one score per unit
head(efficiency(m4, type = "transient"), 3)    # one per observation
```

The unit effect enters none of the three scores, which is the point: in the
time-invariant panel model every persistent difference between units — funding
structure, soil quality, vintage of equipment — has nowhere to go but into
inefficiency.

The combination `mu_i - eta_i` is determined sharply by the data, but how it
divides between the two depends on which of them has the wider spread: where the
unit effect dominates it is recovered well and the persistent term is not, and
where it is small the reverse holds. That is a property of the model rather than
of the sampler, so a prior sensitivity check on `sigma_mu` and the persistent
inefficiency parameter is worth running before reading much into the split. The
frontier and the transient scores are unaffected. See the `four-component`
vignette.

## Roadmap

The planned extensions, roughly in order:

1. **Inefficiency determinants** — a Battese and Coelli (1995) style regression
   in the mean or scale of the inefficiency distribution, estimated jointly with
   the frontier rather than in an inconsistent second step.
2. **Heteroskedasticity** in both error components, which matters whenever unit
   size is widely dispersed.
3. **Latent class / regime switching**, with class membership driven by
   covariates. Gibbs handles the discrete membership indicator naturally, where
   Hamiltonian Monte Carlo would not.
4. **The closed skew normal likelihood** of Colombi et al. (2014), which would
   make information criteria available for the four-component model. It needs
   normal distribution functions of dimension `T_i + 1`.
5. **LOOIC**, once the Pareto smoothed importance sampling it needs has had its
   own testing pass.

Each of these is a block of the specification rather than an argument to an
estimator, which is why the package is organised around a model object.

Note that the pointwise log-likelihood, and therefore every criterion, is
unavailable for panel models: the units share one inefficiency term, so
integrating it out couples the observations that belong to the same unit.

## Validation

The exponential, time-invariant panel specification reproduces the sampler
distributed by Justin Tobias for exercise 14.13 of Koop, Poirier and Tobias
(2007), *Bayesian Econometric Methods*. The `koop-exercise` vignette translates
both that program and its data-generating script into `bsfa` and compares the
results, which is worth having because the data-generating process is known.

The correspondence is exact. The mean of the truncated normal full conditional
for `z_i` carries the `-1/(T h mu_z)` shift contributed by the exponential
prior, which in the half-normal case is replaced by an additional `1/sigma_u^2`
term in the precision; that one line is the whole difference between the two
specifications inside the sampler.

The closed-form log-likelihoods are checked against direct numerical
integration of the composed-error density in the test suite, for both
distributions and for both orientations of the frontier.

## References

Aigner, D., Lovell, C. A. K., & Schmidt, P. (1977). Formulation and estimation
of stochastic frontier production function models. *Journal of Econometrics*,
6(1), 21–37.

Battese, G. E., & Coelli, T. J. (1995). A model for technical inefficiency
effects in a stochastic frontier production function for panel data.
*Empirical Economics*, 20(2), 325–332.

Colombi, R., Kumbhakar, S. C., Martini, G., & Vittadini, G. (2014). Closed skew
normal distribution and efficiency analysis. *Journal of Productivity Analysis*,
42(2), 123–136.

Jondrow, J., Lovell, C. A. K., Materov, I. S., & Schmidt, P. (1982). On the
estimation of technical inefficiency in the stochastic frontier production
function model. *Journal of Econometrics*, 19(2–3), 233–238.

Koop, G. (2003). *Bayesian Econometrics*. Chichester: Wiley.

Kumbhakar, S. C., Lien, G., & Hjalmarsson, L. (2014). Technical efficiency in
competing panel data models: A study of Norwegian grain farming. *Journal of
Productivity Analysis*, 41(2), 321–337.

Pitt, M. M., & Lee, L.-F. (1981). The measurement and sources of technical
inefficiency in the Indonesian weaving industry. *Journal of Development
Economics*, 9(1), 43–64.

Robert, C. P. (1995). Simulation of truncated normal variables. *Statistics and
Computing*, 5(2), 121–125.

van den Broeck, J., Koop, G., Osiewalski, J., & Steel, M. F. J. (1994).
Stochastic frontier models: A Bayesian perspective. *Journal of Econometrics*,
61(2), 273–303.

## Citation

```r
citation("bsfa")
```

Released versions are archived on Zenodo, which mints a DOI for each release
and a concept DOI that always resolves to the latest one. Please cite the
software alongside van den Broeck, Koop, Osiewalski and Steel (1994), whose
sampler it follows.

## License

GPL (>= 2)
