---
title: "Bayesian stochastic frontier analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian stochastic frontier analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r setup}
library(bsfa)
set.seed(1234)
```

## The model

A stochastic frontier model splits the disturbance of a regression into a
symmetric noise term and a one-sided term that measures the distance to a
frontier. For a production frontier,

$$y_i = x_i' \beta - u_i + v_i, \qquad v_i \sim N(0, \sigma_v^2), \quad u_i \ge 0,$$

so that $x_i'\beta$ traces the maximum output attainable with the inputs $x_i$
and $r_i = \exp(-u_i)$ is the efficiency with which unit $i$ reaches it. For a
cost frontier the sign of $u_i$ is reversed and the frontier becomes a lower
envelope.

The package supports two distributions for $u_i$: exponential with rate
$\lambda$, and half-normal with scale $\sigma_u$.

## Building a model

A model is assembled step by step rather than in a single call. The constructor
fixes the data, the frontier and the MCMC settings; the `add_*` functions attach
the remaining blocks; `draw_posterior()` runs the sampler.

```{r}
d <- sim_sf(n = 500, beta = c(1, 0.5, 0.3), sigma_v = 0.2, par_u = 4)

model <- create_sfmodel_exp(y ~ x1 + x2, data = d,
                            iterations = 4000, burnin = 1000)
model
```

Printing the object reports what is still missing. That seam is worth having:
these samplers take a while on a real panel, and it is cheaper to check the
specification here than to discover a mistake after the run.

```{r}
model <- add_priors(model)
model <- add_initial_values(model)
model <- add_seed(model, 5555)
model
```

The distribution of the inefficiency term is carried by the class of the object,
`sfmodel_exp` here and `sfmodel_hn` for the half-normal alternative, so
`add_priors()` and `draw_posterior()` dispatch on it. The next section shows why
that matters more than it might appear.

## Choosing a prior

The inefficiency parameter is hard to reason about directly, so its prior is
anchored on `r_star`, the efficiency score one considers equally likely to be
exceeded as not, a priori.

For the exponential model the mapping is exact. Placing a gamma prior with shape
1 and rate $c = -\log(r^*)$ on $\lambda$ implies the marginal prior density

$$p(u) = \int_0^\infty \lambda e^{-\lambda u} \, c\, e^{-c \lambda} \, d\lambda
        = \frac{c}{(u + c)^2},$$

whose median is exactly $c$, so that the prior median of $\exp(-u)$ is $r^*$.
A quick check:

```{r}
m <- add_priors(create_sfmodel_exp(y ~ x1 + x2, data = d),
                lambda = list(r_star = 0.75))
lambda <- rgamma(1e5, shape = m$priors$shape_u, rate = m$priors$rate_u)
median(exp(-rexp(length(lambda), rate = lambda)))
```

No such closed form is available for the half-normal model, where the anchor is
matched only in expectation: the prior mean of $\sigma_u^2$ is set to the value
that would place the median of $u$ at $-\log(r^*)$. The two models therefore
need different elicitations of the same quantity, which is exactly what the
separate `add_priors()` methods provide — and why the argument is named after
the parameter it governs, `lambda` for one model and `sigma_u` for the other.

## A cross-sectional example

```{r}
est <- draw_posterior(model)
summary(est)
```

The simulated data used $\beta = (1, 0.5, 0.3)$, $\sigma_v = 0.2$ and
$\lambda = 4$. Note that the slopes are pinned down far more sharply than the
intercept, which has to be separated from the mean of the one-sided term using
only the asymmetry of the residuals. That is a feature of the model rather than
of the sampler, and it is worth keeping in mind whenever the level of the
frontier is itself of interest.

Efficiency scores, with their posterior uncertainty:

```{r}
eff <- efficiency(est)
head(eff)
```

```{r}
plot(attr(d, "efficiency"), eff$mean,
     xlab = "Simulated efficiency", ylab = "Posterior mean",
     pch = 16, col = grey(0.3, 0.5))
abline(0, 1)
```

The scatter around the 45-degree line is genuine, not a defect: with one
observation per unit there is very little information about any individual
$u_i$, and the posterior shrinks each score towards the prior. The width of the
intervals in `eff` reports exactly how much.

```{r}
summary(eff[["95%"]] - eff[["5%"]])
```

## Panel data

Supplying `id` gives each unit a single inefficiency term held fixed over its
observations, the time-invariant model of Pitt and Lee (1981). Repeated
observations sharpen the individual scores considerably.

```{r}
dp <- sim_sf(n = 100, beta = c(1, 0.5, 0.3), sigma_v = 0.2, par_u = 4,
             n_time = 8)

estp <- create_sfmodel_exp(y ~ x1 + x2, data = dp, id = "id",
                           iterations = 4000, burnin = 1000)
estp <- draw_posterior(add_priors(estp))

effp <- efficiency(estp)
summary(effp[["95%"]] - effp[["5%"]])
cor(effp$mean, attr(dp, "efficiency"))
```

The price is a strong assumption. Since $u_i$ is the only unit-specific term in
the model, every persistent difference between units — technology, market,
business model — is booked as inefficiency. Separating those two requires a
model with both a unit effect and a persistent inefficiency component, which is
on the roadmap rather than in the package today.

## Checking convergence

`add_initial_values(method = "prior")` draws the starting values from the prior
instead of anchoring them on a least squares fit, which is the more useful
choice when running several chains: the starting points are then dispersed
rather than identical.

```{r}
base <- add_priors(create_sfmodel_exp(y ~ x1 + x2, data = d,
                                      iterations = 2000, burnin = 1000))

chains <- lapply(1:3, function(i) {
  m <- add_seed(add_initial_values(base, method = "prior"), i)
  colMeans(draw_posterior(m, keep_u = FALSE)$draws$beta)
})
do.call(rbind, chains)
```

## References

Jondrow, J., Lovell, C. A. K., Materov, I. S., & Schmidt, P. (1982). On the
estimation of technical inefficiency in the stochastic frontier production
function model. *Journal of Econometrics*, 19(2–3), 233–238.

Koop, G. (2003). *Bayesian Econometrics*. Chichester: Wiley.

Pitt, M. M., & Lee, L.-F. (1981). The measurement and sources of technical
inefficiency in the Indonesian weaving industry. *Journal of Development
Economics*, 9(1), 43–64.

van den Broeck, J., Koop, G., Osiewalski, J., & Steel, M. F. J. (1994).
Stochastic frontier models: A Bayesian perspective. *Journal of Econometrics*,
61(2), 273–303.
