---
title: "Bayesian stochastic frontier analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian stochastic frontier analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
```

```{r setup}
library(bsfa)
set.seed(1234)
```

## The model

A stochastic frontier model splits the disturbance of a regression into a
symmetric noise term and a one-sided term that measures the distance to a
frontier. For a production frontier,

$$y_i = x_i' \beta - u_i + v_i, \qquad v_i \sim N(0, \sigma_v^2), \quad u_i \ge 0,$$

so that $x_i'\beta$ traces the maximum output attainable with the inputs $x_i$
and $r_i = \exp(-u_i)$ is the efficiency with which unit $i$ reaches it. For a
cost frontier the sign of $u_i$ is reversed and the frontier becomes a lower
envelope.

The package supports two distributions for $u_i$: exponential with rate
$\lambda$, and half-normal with scale $\sigma_u$.

## Building a model

A model is assembled step by step rather than in a single call. The constructor
fixes the data, the frontier and the MCMC settings; the `add_*` functions attach
the remaining blocks and then the posterior draws.

```{r}
d <- sim_sf(n = 500, beta = c(1, 0.5, 0.3), sigma_v = 0.2, par_u = 4)

model <- create_sfmodel_exp(y ~ x1 + x2, data = d,
                            iterations = 4000, burnin = 1000)
model
```

Printing the object reports what is still missing. That seam is worth having:
these samplers take a while on a real panel, and it is cheaper to check the
specification here than to discover a mistake after the run.

```{r}
model <- add_priors(model)
model <- add_initial_values(model)
model <- add_seed(model, 5555)
model
```

`add_initial_values()` also draws a seed if the model does not have one, so a
model is reproducible whether or not `add_seed()` is used. The sampler sets that
seed and puts the generator back afterwards, which means a `set.seed()` call
placed between these steps and the simulation does not change the result.

The distribution of the inefficiency term is carried by the class of the object,
`sfmodel_exp` here and `sfmodel_hn` for the half-normal alternative, so
`add_priors()` and `add_posterior_coefficients()` dispatch on it. The next
section shows why that matters more than it might appear.

## Choosing a prior

The inefficiency parameter is hard to reason about directly, so its prior is
anchored on `r_star`, the efficiency score one considers equally likely to be
exceeded as not, a priori.

For the exponential model the mapping is exact. Placing a gamma prior with shape
1 and rate $c = -\log(r^*)$ on $\lambda$ implies the marginal prior density

$$p(u) = \int_0^\infty \lambda e^{-\lambda u} \, c\, e^{-c \lambda} \, d\lambda
        = \frac{c}{(u + c)^2},$$

whose median is exactly $c$, so that the prior median of $\exp(-u)$ is $r^*$.
A quick check:

```{r}
m <- add_priors(create_sfmodel_exp(y ~ x1 + x2, data = d),
                lambda = list(r_star = 0.75))
lambda <- rgamma(1e5, shape = m$priors$shape_u, rate = m$priors$rate_u)
median(exp(-rexp(length(lambda), rate = lambda)))
```

The half-normal model reaches the same place by a different route. Its prior
sits on $\sigma_u^2$ and is inverse gamma, so $u$ is marginally half $t$ on
$2a$ degrees of freedom, and the rate is set from the median of that. The
anchor is exact there too, and at any shape. The two models therefore need
different elicitations of the same quantity, which is exactly what the
separate `add_priors()` methods provide — and why the argument is named after
the parameter it governs, `lambda` for one model and `sigma_u` for the other.

## Drawing from the posterior

```{r}
model <- add_posterior_coefficients(model)
summary(model)
```

The simulated data used $\beta = (1, 0.5, 0.3)$, $\sigma_v = 0.2$ and
$\lambda = 4$. Note that the slopes are pinned down far more sharply than the
intercept, which has to be separated from the mean of the one-sided term using
only the asymmetry of the residuals. That is a feature of the model rather than
of the sampler, and it is worth keeping in mind whenever the level of the
frontier is itself of interest.

The `ESS` column is the effective sample size of each block: the number of
independent draws the chain is worth, which is what the width of a credible
band should be read against. The draws are kept as `coda` objects, one block
per parameter, so anything that reads `mcmc` objects works on them directly if
more is wanted:

```{r}
coda::autocorr.diag(model$posterior$lambda$coeffs, lags = c(1, 5, 10))
```

```{r, fig.height = 5}
plot(model, type = "trace")
```

## Is there inefficiency to find?

Before reading the scores, it is worth asking whether the sample says anything
about the one-sided term. A production frontier skews the composed error to the
left, and if the least squares residuals carry no such skew then there is no
evidence of inefficiency in the data.

```{r}
skewness_test(model)
```

Maximum likelihood answers this loudly, by collapsing onto least squares. A
posterior cannot: the prior holds the inefficiency term away from zero, so the
sampler returns plausible looking scores whether or not the data support them.
A verdict of `weak` or `wrong` does not mean the model should be abandoned, but
it does mean the scores below are largely the prior, and should be reported as
such.

The test is a proxy, and a test, with an error rate of its own. The quantity it
stands for can be measured directly, by moving the anchor and seeing how far
the scores follow:

```{r}
prior_sensitivity(model, r_star = c(0.6, 0.75, 0.9))
```

The frontier coefficients hold still while the scores are free to move, which
is the pattern to expect: what the data determine sharply is the frontier, and
what they determine least is the distance to it.

## Efficiency scores

```{r}
eff <- efficiency(model)
head(eff)
```

```{r}
plot(model, type = "efficiency")
```

The left panel is the summary usually reported. The right one is the reason to
look further: with one observation per unit there is very little information
about any individual $u_i$, and the bands are wide enough to cover much of the
range the histogram spreads over. That uncertainty is real, and it is what the
maximum likelihood route cannot report, since the Jondrow et al. (1982)
conditional mean is a point predictor of an unobserved quantity.

```{r}
summary(eff[["95%"]] - eff[["5%"]])
```

## Comparing specifications

`add_posterior_loglik()` evaluates the pointwise log-likelihood with the
inefficiency term integrated out, and `selection_criteria()` turns those draws
into information criteria.

```{r}
fit <- function(model) {
  model <- add_posterior_coefficients(add_priors(model))
  add_posterior_loglik(model)
}

m_exp <- fit(create_sfmodel_exp(y ~ x1 + x2, data = d,
                                iterations = 4000, burnin = 1000))
m_hn <- fit(create_sfmodel_hn(y ~ x1 + x2, data = d,
                              iterations = 4000, burnin = 1000))

selection_criteria(m_exp)
selection_criteria(m_hn)
```

The data were generated with exponential inefficiency, and the criteria prefer
that specification. They are not always so decisive: the two distributions are
hard to tell apart when inefficiency is small relative to noise, which is the
usual situation.

## Which regressors belong in the frontier

Creating the model with `varsel` places the frontier coefficients under a
variable selection algorithm, which draws an inclusion indicator for each
candidate regressor in every sweep rather than leaving the choice to be argued
out of a table of credible bands afterwards.

```{r}
d$z1 <- rnorm(nrow(d))   # a regressor the frontier has no use for

sel <- create_sfmodel_exp(y ~ x1 + x2 + z1, data = d, varsel = "ssvs",
                          iterations = 4000, burnin = 1000)
sel <- add_priors(sel, varsel = list(semiautomatic = c(0.1, 10)))
sel <- add_posterior_coefficients(add_seed(sel, 1234))

summary(sel)$coefficients
```

The `PIP` column is the share of draws in which the regressor was in the
frontier: the two that generated the data are kept throughout, the one that did
not is dropped in most sweeps. The mean and the band on the same row are
averages over both states of the indicator, so this is model averaging rather
than model choice, and the efficiency scores come out averaged over the
frontiers the selection admits.

`"ssvs"` is the stochastic search variable selection of George, Sun and Ni
(2008), which tightens the prior of an excluded coefficient, and `"bvs"` is the
Bayesian variable selection of Korobilis (2013), which switches the regressor
itself off so that an excluded coefficient is stored as an exact zero. The two
differ in what they need from the prior and in how they mix; see the
`variable-selection` vignette.

## Panel data

Supplying `id` gives each unit a single inefficiency term held fixed over its
observations, the time-invariant model of Pitt and Lee (1981). Repeated
observations sharpen the individual scores considerably.

```{r}
dp <- sim_sf(n = 100, beta = c(1, 0.5, 0.3), sigma_v = 0.2, par_u = 4,
             n_time = 8)

estp <- create_sfmodel_exp(y ~ x1 + x2, data = dp, id = "id",
                           iterations = 4000, burnin = 1000)
estp <- add_posterior_coefficients(add_priors(estp))

effp <- efficiency(estp)
summary(effp[["95%"]] - effp[["5%"]])
cor(effp$mean, attr(dp, "efficiency"))
```

The price is a strong assumption. Since $u_i$ is the only unit-specific term in
the model, every persistent difference between units — technology, market,
business model — is booked as inefficiency. Separating those two requires a
model with both a unit effect and a persistent inefficiency component, which is
on the roadmap rather than in the package today.

Note also that `add_posterior_loglik()` refuses a panel model. The units share
one inefficiency term, so integrating it out couples the observations that
belong to the same unit and the likelihood no longer factorises over them. The
information criteria are therefore unavailable here, and a panel specification
cannot be compared with a cross-sectional one this way.

## Checking convergence

`add_initial_values(method = "prior")` draws the starting values from the prior
instead of anchoring them on a least squares fit, which is the more useful
choice when running several chains: the starting points are then dispersed
rather than identical.

```{r}
base <- add_priors(create_sfmodel_exp(y ~ x1 + x2, data = d,
                                      iterations = 2000, burnin = 1000))

chains <- lapply(1:3, function(i) {
  m <- add_seed(add_initial_values(base, method = "prior"), i)
  colMeans(add_posterior_coefficients(m, keep_u = FALSE)$posterior$beta$coeffs)
})
do.call(rbind, chains)
```

## References

George, E. I., Sun, D., & Ni, S. (2008). Bayesian stochastic search for VAR
model restrictions. *Journal of Econometrics*, 142(1), 553–580.

Jondrow, J., Lovell, C. A. K., Materov, I. S., & Schmidt, P. (1982). On the
estimation of technical inefficiency in the stochastic frontier production
function model. *Journal of Econometrics*, 19(2–3), 233–238.

Koop, G. (2003). *Bayesian Econometrics*. Chichester: Wiley.

Korobilis, D. (2013). VAR forecasting using Bayesian variable selection.
*Journal of Applied Econometrics*, 28(2), 204–230.

Pitt, M. M., & Lee, L.-F. (1981). The measurement and sources of technical
inefficiency in the Indonesian weaving industry. *Journal of Development
Economics*, 9(1), 43–64.

van den Broeck, J., Koop, G., Osiewalski, J., & Steel, M. F. J. (1994).
Stochastic frontier models: A Bayesian perspective. *Journal of Econometrics*,
61(2), 273–303.
