---
title: "Determinants of inefficiency"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Determinants of inefficiency}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library("bsfa")
```

A frontier model without determinants answers how inefficient each unit is.
It does not answer why, and a ranking on its own rarely tells anyone what to
change. Determinants close that gap by letting covariates enter the
distribution of the inefficiency term itself, so that the posterior says not
only that a unit is far from the frontier but what it is about the unit that
puts it there.

There are two places a covariate can enter, and they say different things.

# Where the covariates go

## The scale

The first is the scale. Write the inefficiency term's scale parameter as the
baseline times a factor,

$$\sigma_{u,i}^2 = \sigma_u^2 \exp(z_i' \gamma),$$

for a half-normal or truncated normal term, and the same factor on the mean
of an exponential one. A positive $\gamma$ means the term is more spread out
where the determinant is large, and since the term is one-sided, more spread
out means larger on average: the distribution stays anchored at zero and
stretches away from it. This is the heteroskedastic specification of Caudill,
Ford and Gropper (1995), and it is what the applied four-component literature
reaches for.

The factor multiplies the baseline rather than replacing it, which has two
consequences worth knowing. A coefficient of zero returns the model without
determinants exactly, so nothing is given up by allowing them. And the
baseline keeps the meaning it has in that model, so the `r_star` elicitation
of `add_priors()` still anchors the prior median efficiency of a unit whose
determinants are all zero. That is why the set carries no intercept: the
baseline already is one, and a column of ones beside it would be the same
quantity written twice.

## The pre-truncation mean

The second is the mean, and only the truncated normal family has one. A
half-normal or exponential term is pinned at zero, and the only thing a
covariate can do to it is change how far it spreads. A truncated normal,

$$u_i \sim N^+(z_i' \delta, \sigma_u^2),$$

is a normal distribution before truncation whose centre can be moved, so a
covariate can shift the whole distribution rather than only stretch it. With
nothing but an intercept this is Stevenson (1980); with covariates it is
Battese and Coelli (1995); with `scale_u` as well it is Wang (2002), where a
covariate can change the mean and the spread in opposite directions and the
effect on efficiency need not be monotone.

Which to use is a modelling question and not a technical one. The scale is
the more conservative choice, since it cannot make the distribution
anything other than one-sided at zero. The mean is the more direct one, since
$\delta$ reads as the effect of the covariate on inefficiency in the units of
the response.

# A scale determinant

Simulated data where inefficiency is more variable, and so larger on average,
when `z1` is large:

```{r}
set.seed(1234)

n <- 500
z1 <- rnorm(n)
x1 <- rnorm(n)
x2 <- rnorm(n)

u <- abs(rnorm(n, mean = 0, sd = 0.3 * sqrt(exp(0.8 * z1))))
y <- 1 + 0.5 * x1 + 0.3 * x2 - u + rnorm(n, sd = 0.2)

d <- data.frame(y, x1, x2, z1)
```

The determinants are a one-sided formula, given to the constructor, exactly
as a variable selection algorithm is:

```{r}
het <- create_sfmodel_hn(y ~ x1 + x2,
                         data = d,
                         scale_u = ~ z1,
                         iterations = 4000,
                         burnin = 2000)

het <- add_priors(het)

het <- add_posterior_coefficients(add_seed(het, 1234))

summary(het)
```

The row named `z1` is $\gamma$, and the summary says which term it explains.
Its band excludes zero, which is the statement that the determinant matters;
`sigma_u` is what the scale would be at `z1 = 0`.

Read the size of $\gamma$ on the log scale it lives on. A coefficient of 0.8
means the variance of the inefficiency term is $e^{0.8} \approx 2.2$ times
larger for a unit one standard deviation above the mean of `z1`, and its
standard deviation $e^{0.4} \approx 1.5$ times larger. It does not mean
efficiency falls by 0.8 of anything.

## The acceptance rate

The summary prints something the other models do not have. Neither $\gamma$
nor $\delta$ has a conjugate full conditional, so they are not drawn in the
way the rest of the sampler draws its blocks; each is updated by a random
walk Metropolis step inside the sweep. The proposal is tuned during burn-in
towards an acceptance rate of 0.234 and then left alone, so the retained
draws come from a kernel with the right invariant distribution.

The rate is reported because a block that rarely moves produces draws that
look like a tight posterior and are nothing of the sort. A rate near the
target and an effective sample size in the hundreds mean the block is
working; a rate near zero means the chain sat still, and the band on that row
should not be believed. Give the run a longer burn-in if that happens, so
that the adaptation has more to work with.

# A mean determinant

The same covariate, now shifting the centre of a truncated normal:

```{r}
set.seed(4321)

m_u <- 0.2 + 0.45 * z1
ut <- numeric(n)
for (i in seq_len(n)) {
  repeat {
    v <- rnorm(1, mean = m_u[i], sd = 0.25)
    if (v > 0) {
      ut[i] <- v
      break
    }
  }
}

y2 <- 1 + 0.5 * x1 + 0.3 * x2 - ut + rnorm(n, sd = 0.2)
d2 <- data.frame(y = y2, x1, x2, z1)
```

```{r}
bc <- create_sfmodel_tn(y ~ x1 + x2,
                        data = d2,
                        mean_u = ~ z1,
                        iterations = 4000,
                        burnin = 2000)

bc <- add_priors(bc)

bc <- add_posterior_coefficients(add_seed(bc, 4321))

summary(bc)
```

Two rows are now named `(Intercept)`: one in the frontier and one in
`mean_u`. The second is the pre-truncation mean where `z1` is zero, and it is
the parameter that makes the truncated normal more than a half-normal. The
`z1` row beside it is the effect of the covariate on that mean, in the units
of the response.

The frontier intercept is the coefficient to read with most care here. It is
identified only jointly with the mean of the inefficiency term, since both
shift the fitted line by a constant, and its effective sample size will
usually be the smallest in the table for that reason.

# Two terms, two sets

In the four-component model each one-sided term takes its own set, and they
sit at different levels. The persistent term is one per unit, so a
determinant of it has to be a property of the unit; the transient term is one
per observation, so a determinant of it may vary within one.

```{r}
set.seed(11)

n_units <- 80
n_time <- 6
id <- rep(seq_len(n_units), each = n_time)

ze <- rep(rnorm(n_units), each = n_time)    # unit level
zu <- rnorm(n_units * n_time)               # varies within a unit
xp <- rnorm(n_units * n_time)

mu_i <- rep(rnorm(n_units, sd = 0.3), each = n_time)
eta_i <- rep(abs(rnorm(n_units, sd = 0.25)), each = n_time) * sqrt(exp(0.9 * ze))
u_it <- abs(rnorm(n_units * n_time, sd = 0.2 * sqrt(exp(0.6 * zu))))

yp <- 1 + 0.5 * xp + mu_i - eta_i - u_it + rnorm(n_units * n_time, sd = 0.15)
dp <- data.frame(y = yp, x1 = xp, ze, zu, id)
```

```{r}
four <- create_sfmodel4_hn(y ~ x1,
                           data = dp,
                           id = "id",
                           scale_eta = ~ ze,
                           scale_u = ~ zu,
                           iterations = 4000,
                           burnin = 2000)

four <- add_posterior_coefficients(add_seed(add_priors(four), 11))

summary(four)
```

Giving a within-unit variable to the persistent term is refused rather than
averaged behind your back, because the two are different models and only you
know which was meant:

```{r, error = TRUE}
create_sfmodel4_hn(y ~ x1, data = dp, id = "id", scale_eta = ~ zu)
```

This division is the point of the four-component model carried into the
determinants. A business model index measured every quarter belongs on the
transient term; its variability over the whole sample, one number per unit,
belongs on the persistent one. That pairing is what Badunenko, Kumbhakar and
Lozano-Vivas (2021) use to separate a bank's short-run adjustments from its
long-run strategy.

Note how much wider the band on `ze` is than the band on `zu`. That is the
identification problem the four-component model already has, reappearing
where you would expect it: the persistent inefficiency term is told apart
from the unit effect only by its sign restriction, so a covariate explaining
it is estimated from whatever the data have left to say once that split is
made. The transient term is identified within a unit and its determinant is
correspondingly sharper. `prior_sensitivity()` is worth running before
reading much into a persistent-side coefficient.

# What determinants cost

The pointwise log-likelihood is not available for a model carrying
determinants, and with it `selection_criteria()` is not either:

```{r, error = TRUE}
add_posterior_loglik(het)
```

The closed form that integrates the one-sided term out of the composed error
assumes one scale shared by every observation, and with determinants there is
a different scale at each. The same applies to the truncated normal whatever
its determinants, whose composed error is a closed skew normal rather than
the skew normal the half-normal gives. Rather than return a number that looks
like a log-likelihood and is not one, the case is refused.

Everything built on the augmented draws is unaffected. `efficiency()`,
`plot()` and `prior_sensitivity()` work as they do for any other model, since
they read the inefficiency terms the sampler produced rather than a
distribution assumed for them:

```{r}
head(efficiency(het), 3)
```

# References

Badunenko, O., Kumbhakar, S. C., & Lozano-Vivas, A. (2021). Achieving a
sustainable cost-efficient business model in banking: The case of European
commercial banks. *European Journal of Operational Research*, 293(2),
773--785.

Battese, G. E., & Coelli, T. J. (1995). A model for technical inefficiency
effects in a stochastic frontier production function for panel data.
*Empirical Economics*, 20(2), 325--332.

Caudill, S. B., Ford, J. M., & Gropper, D. M. (1995). Frontier estimation and
firm-specific inefficiency measures in the presence of heteroscedasticity.
*Journal of Business & Economic Statistics*, 13(1), 105--111.

Roberts, G. O., & Rosenthal, J. S. (2009). Examples of adaptive MCMC.
*Journal of Computational and Graphical Statistics*, 18(2), 349--367.

Stevenson, R. E. (1980). Likelihood functions for generalized stochastic
frontier estimation. *Journal of Econometrics*, 13(1), 57--66.

Wang, H.-J. (2002). Heteroscedasticity and non-monotonic efficiency effects
of a stochastic frontier model. *Journal of Productivity Analysis*, 18(3),
241--253.
