---
title: "The four-component model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The four-component model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
```

```{r setup}
library(bsfa)
set.seed(1234)
```

## Why four components

The time-invariant panel model of `create_sfmodel_exp()` gives each unit a
single inefficiency term held fixed over its observations. Two things follow,
and both are usually wrong.

Every persistent difference between units is booked as inefficiency. A bank with
an unusual funding structure, a farm on poorer soil, a plant with older
equipment: all of it lands in $u_i$ and depresses the score, whether or not it
has anything to do with how well the unit is run.

And nothing can move. A model whose inefficiency term is constant within a unit
cannot answer whether efficiency improved over the sample, which is often the
question that motivated the exercise.

The four-component model of Kumbhakar, Lien and Hjalmarsson (2014), Colombi et
al. (2014) and Tsionas and Kumbhakar (2014) splits the disturbance into

$$y_{it} = x_{it}' \beta + \mu_i - \eta_i - u_{it} + v_{it},$$

with $\mu_i \sim N(0, \sigma_\mu^2)$ an unrestricted unit effect, $\eta_i \ge 0$
persistent inefficiency, $v_{it}$ noise and $u_{it} \ge 0$ transient
inefficiency. The unit effect absorbs heterogeneity that is not inefficiency;
the two one-sided terms separate what does not move from what does.

## Fitting one

The chain is the one the rest of the package uses. `id` is required here, since
the components cannot be separated without repeated observations.

```{r}
d <- sim_sf4(n = 120, n_time = 8, beta = c(1, 0.5, 0.3),
             sigma_v = 0.2, sigma_mu = 0.15, par_eta = 4, par_u = 6)

model <- create_sfmodel4_exp(y ~ x1 + x2, data = d, id = "id",
                             iterations = 4000, burnin = 1000)
model <- add_priors(model,
                    lambda_eta = list(r_star = 0.9),
                    lambda_u = list(r_star = 0.9))
model <- add_seed(add_initial_values(model), 99)
model <- add_posterior_coefficients(model)

summary(model)
```

The two one-sided terms are elicited separately, each from its own prior median
efficiency. Note that those anchors apply per component: with `r_star = 0.9` for
each, the prior median of the overall efficiency $\exp(-\eta_i - u_{it})$ is
near 0.81.

## Three efficiency scores

This is what the model is for.

```{r}
pers <- efficiency(model, type = "persistent")
tran <- efficiency(model, type = "transient")
over <- efficiency(model, type = "overall")

c(persistent = nrow(pers), transient = nrow(tran), overall = nrow(over))
```

Persistent efficiency is one score per unit and transient efficiency one per
observation, so they answer different questions. The first is what an investment
or a policy would have to change; the second is what management can move within
the year.

```{r}
plot(exp(-attr(d, "eta")), pers$mean,
     xlab = "Simulated persistent efficiency", ylab = "Posterior mean",
     pch = 16, col = grey(0.3, 0.6))
abline(0, 1)
```

The unit effect enters none of the three scores. That is the point: in the
two-component model it has nowhere to go but into inefficiency.

```{r}
two <- create_sfmodel_exp(y ~ x1 + x2, data = d, id = "id",
                          iterations = 4000, burnin = 1000)
two <- add_posterior_coefficients(add_priors(two))

c(four_component_persistent = mean(pers$mean),
  two_component = mean(efficiency(two)$mean))
```

## What is identified, and what is not

$\mu_i$ and $\eta_i$ are both constant within a unit and are told apart only by
the sign restriction on $\eta$ and by the distributional assumptions. The
combination $\mu_i - \eta_i$ is determined sharply; how it divides between the
two depends on which has the wider spread.

```{r}
mu_hat <- colMeans(model$posterior$mu$coeffs)
eta_hat <- colMeans(model$posterior$eta$coeffs)

c(combination = cor(mu_hat - eta_hat, attr(d, "mu") - attr(d, "eta")),
  unit_effect = cor(mu_hat, attr(d, "mu")),
  persistent = cor(eta_hat, attr(d, "eta")))
```

Repeating the exercise with a unit effect twice as wide and a persistent term
half as large reverses which of the two comes out well, while leaving the
combination untouched:

```{r}
d2 <- sim_sf4(n = 120, n_time = 8, beta = c(1, 0.5, 0.3),
              sigma_v = 0.2, sigma_mu = 0.3, par_eta = 8, par_u = 6)
m2 <- create_sfmodel4_exp(y ~ x1 + x2, data = d2, id = "id",
                          iterations = 4000, burnin = 1000)
m2 <- add_posterior_coefficients(add_priors(m2))

mu2 <- colMeans(m2$posterior$mu$coeffs)
eta2 <- colMeans(m2$posterior$eta$coeffs)

c(combination = cor(mu2 - eta2, attr(d2, "mu") - attr(d2, "eta")),
  unit_effect = cor(mu2, attr(d2, "mu")),
  persistent = cor(eta2, attr(d2, "eta")))
```

This is a property of the model rather than of the sampler, and it is the reason
to run a prior sensitivity check on `sigma_mu` and on the persistent
inefficiency parameter before reading much into the split. The frontier itself,
and the transient scores, are unaffected.

## Convergence

The same trade-off shows up in the chain: $\mu_i$ and $\eta_i$ can swap against
each other from sweep to sweep. Starting several chains from the prior is worth
the time here.

```{r}
base <- add_priors(create_sfmodel4_exp(y ~ x1 + x2, data = d, id = "id",
                                       iterations = 2000, burnin = 1000))

chains <- lapply(1:3, function(i) {
  m <- add_seed(add_initial_values(base, method = "prior"), i)
  m <- add_posterior_coefficients(m, keep_u = FALSE)
  c(sigma_mu = mean(m$posterior$sigma_mu$coeffs),
    lambda_eta = mean(m$posterior$lambda_eta$coeffs),
    lambda_u = mean(m$posterior$lambda_u$coeffs))
})
round(do.call(rbind, chains), 3)
```

## What is not available

`add_posterior_loglik()` refuses this model, and with it `selection_criteria()`.
Integrating out the unit effect and the persistent term couples the observations
of a unit, so the likelihood does not factorise over them. A closed form does
exist — the closed skew normal of Colombi et al. (2014) — but it needs normal
distribution functions of dimension $T_i + 1$, which this package does not
implement.

## References

Colombi, R., Kumbhakar, S. C., Martini, G., & Vittadini, G. (2014). Closed skew
normal distribution and efficiency analysis. *Journal of Productivity Analysis*,
42(2), 123–136.

Kumbhakar, S. C., Lien, G., & Hjalmarsson, L. (2014). Technical efficiency in
competing panel data models: A study of Norwegian grain farming. *Journal of
Productivity Analysis*, 41(2), 321–337.

Tsionas, E. G., & Kumbhakar, S. C. (2014). Firm heterogeneity, persistent and
transient technical inefficiency: A generalized true random-effects model.
*Journal of Applied Econometrics*, 29(1), 110–132.
