---
title: "Reproducing the panel frontier exercise of Koop, Poirier and Tobias"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the panel frontier exercise of Koop, Poirier and Tobias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r setup}
library(bsfa)
```

Exercise 14.13 of Koop, Poirier and Tobias (2007), which follows Chapter 7 of
Koop (2003), estimates a panel stochastic frontier model by Gibbs sampling with
data augmentation. The MATLAB programs distributed with the book, `ch14q13dgp.m`
and `ch14q13.m`, generate an artificial panel and then sample from the
posterior. Because the data-generating process is known, the exercise is a
useful external check on this package: the sampler here should reproduce it.

This vignette translates both programs into `bsfa` and compares the results.

## The data-generating process

`ch14q13dgp.m` builds a balanced panel of $N = 100$ individuals observed over
$T = 5$ periods, with an intercept and two regressors drawn from a uniform
distribution:

$$y_{it} = x_{it}'\theta - z_i + \varepsilon_{it}, \qquad
  \varepsilon_{it} \sim N(0, \sigma^2),$$

with $\theta = (1, 0.75, 0.25)'$ and $\sigma = 0.2$, so the error precision is
$h = 25$. The inefficiency term is drawn once per individual and held fixed over
that individual's five periods — the time-invariant specification of Pitt and
Lee (1981).

In the MATLAB the draw is `gamm_rnd(1, 1, .5*vz, .5*vz*muterm)` with `vz = 2`
and `muterm = 1/muz`, which is a gamma with shape `vz/2 = 1` and rate `1/muz`:
an exponential with mean $\mu_z = -\log(0.85)$. The individuals are therefore
about 85 per cent efficient on average by construction.

```{r}
set.seed(14013)

n <- 100
t <- 5
theta <- c(1, 0.75, 0.25)
sigma <- 0.2
mu_z <- -log(0.85)

x <- cbind(x1 = runif(n * t), x2 = runif(n * t))
z <- rexp(n, rate = 1 / mu_z)
id <- rep(seq_len(n), each = t)

y <- as.numeric(cbind(1, x) %*% theta) - z[id] + rnorm(n * t, sd = sigma)

stoch_front <- data.frame(y = y, x, id = id)
head(stoch_front, 3)
```

## The prior

`ch14q13.m` sets an independent normal–gamma prior plus a hierarchical prior on
the inefficiencies. Each of its three blocks maps onto one argument of
`add_priors()`.

The coefficients get a normal prior centred at $(0, 0.5, 0.5)'$, with a variance
of 100 on the intercept and $0.25^2$ on the slopes. `add_priors()` takes the
prior *precision*, so the variances are inverted:

```{r}
v_beta <- diag(c(100, 0.25^2, 0.25^2))
coef_prior <- list(mu = c(0, 0.5, 0.5), v_i = solve(v_beta))
```

Note that this prior is deliberately at odds with the truth: it puts both slopes
at 0.5 with a standard deviation of 0.25, while the data were generated with
0.75 and 0.25.

The error precision gets `v0 = 0` degrees of freedom, which is the improper
limiting prior $p(h) \propto 1/h$. In the shape and rate parameterisation used
here that is a pair of zeros:

```{r}
sigma_prior <- list(shape = 0, rate = 0)
```

The inefficiency prior uses `v_z = 2` and `mu_z = -log(.5)`, that is, a prior
median efficiency of one half. This is the elicitation of van den Broeck, Koop,
Osiewalski and Steel (1994), and `add_priors()` takes the anchor directly:

```{r}
lambda_prior <- list(r_star = 0.5, shape = 1)
```

So the prior expects far more inefficiency than the data contain, which makes
the exercise a reasonable test of whether the likelihood dominates.

## Estimation

`ch14q13.m` discards 25 draws and keeps 250. Those numbers are much too small
for real work, and the book says so, but they are what the reference program
uses:

```{r}
model <- create_sfmodel_exp(y ~ x1 + x2, data = stoch_front, id = "id",
                            iterations = 250, burnin = 25)
model <- add_priors(model,
                    coef = coef_prior,
                    sigma = sigma_prior,
                    lambda = lambda_prior)
model <- add_initial_values(model)
model <- add_seed(model, 14013)
model <- add_posterior_coefficients(model)

summary(model)
```

Supplying `id` is what makes this the Pitt and Lee specification: one
inefficiency term per individual, held fixed over that individual's periods,
exactly as `zdraw` is in the MATLAB loop.

The posterior means sit close to $\theta = (1, 0.75, 0.25)$ despite the prior
pulling both slopes towards 0.5. The error standard deviation is recovered near
0.2, which is the precision of 25 the program prints.

## A longer run

Two hundred and fifty draws is not enough to pin down the tails. With a longer
chain the picture firms up:

```{r}
long <- create_sfmodel_exp(y ~ x1 + x2, data = stoch_front, id = "id",
                           iterations = 20000, burnin = 2000)
long <- add_priors(long,
                   coef = coef_prior,
                   sigma = sigma_prior,
                   lambda = lambda_prior)
long <- add_seed(add_initial_values(long), 14013)
long <- add_posterior_coefficients(long)

round(summary(long)$coefficients, 4)
```

Comparing the two against the values used to simulate the data:

```{r}
truth <- c(theta, sigma, 1 / mu_z)
cbind(truth = truth,
      short = summary(model)$coefficients[, "mean"],
      long = summary(long)$coefficients[, "mean"])
```

The last row is the rate $\lambda$ of the exponential, whose true value is
$1/\mu_z$. It is the least well determined of the parameters: with $N = 100$
individuals there are only a hundred observations on the inefficiency
distribution, however long the panel.

## Figure 7.5

The program closes with a histogram of the posterior means of the efficiencies,
which is Figure 7.5 in Koop (2003).

```{r, fig.width = 7}
plot(long, type = "efficiency")
```

The left panel is that figure. The right one is what the book's program does not
show, and is the more interesting of the two.

Because `bsfa` keeps the augmented draws rather than only their means, the same
object also answers how well any individual score is determined:

```{r}
eff <- efficiency(long)
head(eff)
```

Each interval is wide even with five observations per individual. That is not a
shortcoming of the sampler but the shape of the problem: a single $z_i$ has to
be told apart from five normal errors, so the data leave a good deal of
uncertainty about it. Maximum likelihood faces the same difficulty, and reports
the Jondrow et al. (1982) conditional mean without an interval to say so.

```{r}
plot(exp(-z), eff$mean,
     xlab = "Simulated efficiency", ylab = "Posterior mean",
     pch = 16, col = grey(0.3, 0.6))
abline(0, 1)
```

## How the sampler corresponds to the program

The full conditional for $z_i$ in `ch14q13.m` is a normal truncated at zero,
with

```
zvar  = 1 / (t * h)
zterm = 1 / (t * h * muz)
zmean = mean(x_i) * beta - mean(y_i) - zterm
```

The same conditional is used here. Writing $e_{it} = y_{it} - x_{it}'\beta$, the
precision is $T_i / \sigma_v^2$ and the mean is
$\left(-\sum_t e_{it} / \sigma_v^2 - \lambda\right)$ divided by that precision,
which is the expression above once $\lambda = 1/\mu_z$ and $h = 1/\sigma_v^2$
are substituted. The `zterm` is the contribution of the exponential prior.

The half-normal model of `create_sfmodel_hn()` differs in exactly that place:
there is no shift in the mean, and instead the precision gains a term
$1/\sigma_u^2$. That one line is the whole difference between the two
specifications inside the sampler, and the reason the package splits them by
class is the prior rather than the loop.

## References

Jondrow, J., Lovell, C. A. K., Materov, I. S., & Schmidt, P. (1982). On the
estimation of technical inefficiency in the stochastic frontier production
function model. *Journal of Econometrics*, 19(2–3), 233–238.

Koop, G. (2003). *Bayesian Econometrics*. Chichester: Wiley.

Koop, G., Poirier, D. J., & Tobias, J. L. (2007). *Bayesian Econometric
Methods*. Cambridge: Cambridge University Press.

Pitt, M. M., & Lee, L.-F. (1981). The measurement and sources of technical
inefficiency in the Indonesian weaving industry. *Journal of Development
Economics*, 9(1), 43–64.

van den Broeck, J., Koop, G., Osiewalski, J., & Steel, M. F. J. (1994).
Stochastic frontier models: A Bayesian perspective. *Journal of Econometrics*,
61(2), 273–303.
