---
title: "Variable selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variable selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4)
```

```{r setup}
library(bsfa)
set.seed(1234)
```

## The problem

A frontier is usually written down with every regressor to hand, and the ones
that turn out not to belong are then argued away from a table of credible
bands: this one covers zero, drop it, refit. That reads one coefficient at a
time, and it leaves the efficiency scores conditioned on whichever frontier the
argument happened to settle on, with none of the uncertainty about that choice
anywhere in the answer.

Variable selection puts the choice inside the sampler instead. An inclusion
indicator is drawn for each candidate regressor in every sweep, so the
posterior covers the frontiers as well as the coefficients, and the efficiency
scores come out averaged over them.

`bsfa` offers the two algorithms of the
[bvartools](https://github.com/franzmohr/bvartools) package, the stochastic
search variable selection of George, Sun and Ni (2008) and the Bayesian
variable selection of Korobilis (2013). They differ in what an inclusion
indicator does, which is the subject of most of this vignette, but they are
used in exactly the same way.

## The common interface

Both are switched on with the `varsel` argument of the constructor, and both
take their prior from the `varsel` argument of `add_priors()`, which a model
created with an algorithm requires and every other model refuses:

```r
model <- create_sfmodel_exp(y ~ x1 + x2,
                            data = d,
                            varsel = "ssvs")

model <- add_priors(model,
                    varsel = list(...))
```

The draws of the indicators land in `model$posterior$inclusion`, one column per
candidate, and their means are reported by `summary()` as `PIP`, the posterior
probability that the regressor belongs in the frontier. Everything else about
the model is unchanged: the same constructors, the same priors on the
inefficiency term and the error, the same `efficiency()`, `plot()` and
`selection_criteria()`.

## Data with something to drop

Two of the four regressors below have nothing to do with the frontier. A
selection that works should keep `x1` and `x2` and discard `z1` and `z2`.

```{r}
d <- sim_sf(n = 400, beta = c(1, 0.5, 0.3), sigma_v = 0.2, par_u = 4)
d$z1 <- rnorm(nrow(d))
d$z2 <- rnorm(nrow(d))
```

## Stochastic search variable selection

Under `varsel = "ssvs"` each candidate coefficient carries a mixture of two
normal priors centred on zero: a tight one of standard deviation $\tau_0$
standing for the regressor being absent from the frontier, and a loose one of
$\tau_1 > \tau_0$ standing for its being present. The indicator says which of
the two is in force in that sweep. The regressor never leaves the design
matrix, and an excluded coefficient is held near zero rather than at it.

The two standard deviations can be given directly as `tau`, or derived with
`semiautomatic`, which multiplies the least squares standard error of each
coefficient by the two factors supplied. The second is the one to reach for
when the regressors sit on scales that differ, since the prior then follows the
scale of each coefficient instead of being the same for all of them.

```{r}
ssvs <- create_sfmodel_exp(y ~ x1 + x2 + z1 + z2,
                           data = d,
                           varsel = "ssvs",
                           iterations = 4000,
                           burnin = 1000)

ssvs <- add_priors(ssvs, varsel = list(semiautomatic = c(0.1, 10)))

ssvs <- add_posterior_coefficients(add_seed(ssvs, 1234))

summary(ssvs)
```

The `PIP` column is the share of draws in which the regressor was in the
frontier. The two that generated the data are kept throughout; the two that did
not are excluded in most sweeps.

The intercept is `NA` because it was never a candidate. It is the level of the
frontier rather than the effect of a regressor, so selecting it away would move
every efficiency score instead of dropping a variable, and it is left out by
default. `exclude_intercept = FALSE` puts it in.

## Bayesian variable selection

Under `varsel = "bvs"` the indicator multiplies the regressor rather than its
prior. The frontier is $x_i' \Lambda \beta$ with $\Lambda$ the diagonal matrix
of indicators, so an excluded regressor leaves the likelihood outright and the
coefficient stored for it is an exact zero.

There is no mixture here and so no pair of standard deviations to elicit. The
coefficient keeps the ordinary normal prior of `coef`, and the only thing
`varsel` needs is `inprior`.

```{r}
bvs <- create_sfmodel_exp(y ~ x1 + x2 + z1 + z2,
                          data = d,
                          varsel = "bvs",
                          iterations = 4000,
                          burnin = 1000)

bvs <- add_priors(bvs,
                  coef = list(mu = 0, v_i = 1),
                  varsel = list(inprior = 0.5))

bvs <- add_seed(bvs, 1234)

bvs <- add_posterior_coefficients(bvs)

summary(bvs)
```

The zeros are literal, and they line up with the indicators exactly:

```{r}
beta <- as.matrix(bvs$posterior$beta$coeffs)
inclusion <- as.matrix(bvs$posterior$inclusion$coeffs)

round(colMeans(beta == 0), 3)
round(1 - colMeans(inclusion), 3)
```

This is what `coef()` and `fitted()` then work from: the coefficient a sweep
built its frontier with, zero included, so the posterior mean of a coefficient
is its effect averaged over the frontiers the selection admitted.

## How to read a PIP

A low `PIP` does not mean the coefficient is small; it means the regressor was
usually not in the model. The mean and the credible band on the same row are
averages over both states of the indicator, which is why a coefficient with a
low `PIP` has a posterior concentrated near zero rather than one centred on its
least squares value:

```{r}
round(quantile(as.matrix(ssvs$posterior$beta$coeffs)[, "z1"],
               c(0.05, 0.5, 0.95)), 4)

round(coef(lm(y ~ x1 + x2 + z1 + z2, data = d))[["z1"]], 4)
```

This is model averaging rather than model choice. Nothing has been removed from
the object, and there is no single selected model to report. If a single
frontier is wanted at the end, the usual convention is to keep the regressors
whose `PIP` exceeds one half and refit without the selection, and to say that
that is what was done.

## The prior that Bayesian variable selection needs

The two algorithms part company over what happens to an excluded coefficient,
and this is the practical difference between them.

Under SSVS an excluded coefficient is held at zero by the tight half of its own
prior, so it is always within reach of being included again. Under BVS it is
absent from the likelihood altogether, and the sweep draws it from the prior in
`coef`. The wider that prior, the further the coefficient wanders while it is
out, and a coefficient that has wandered far from anything the data support is
one the likelihood will not admit back.

The effect is not subtle. Below, the same model is fitted at five prior widths.
What matters is not the width itself but the width relative to the scale of the
coefficient, so it is reported as a multiple of the least squares standard
error, which is the scale `add_priors()` measures against too.

```{r}
se <- coef(summary(lm(y ~ x1 + x2 + z1, data = d)))["z1", "Std. Error"]

run <- function(v_i) {
  m <- create_sfmodel_exp(y ~ x1 + x2 + z1,
                          data = d,
                          varsel = "bvs",
                          iterations = 2000,
                          burnin = 1000)

  m <- suppressWarnings(
    add_priors(m, coef = list(mu = 0, v_i = v_i),
               varsel = list(inprior = 0.5)))

  m <- add_posterior_coefficients(add_seed(m, 42))

  z1 <- as.matrix(m$posterior$inclusion$coeffs)[, "z1"]

  data.frame(prior_sd = round(1 / sqrt(v_i), 3),
             multiple_of_se = round((1 / sqrt(v_i)) / se),
             switches = sum(diff(z1) != 0),
             pip_z1 = mean(z1))
}

do.call(rbind, lapply(c(100, 10, 1, 0.1, 0.01), run))
```

Past a couple of hundred times the standard error the indicator never changes
state for the whole run. The posterior inclusion probability of exactly zero
that comes back is not a finding about `z1`; it is a chain that started outside
the model and never got back in. `add_priors()` measures that ratio and warns
above a hundred:

```{r}
bad <- create_sfmodel_exp(y ~ x1 + x2 + z1,
                          data = d,
                          varsel = "bvs",
                          iterations = 100,
                          burnin = 100)

bad <- tryCatch(add_priors(bad, varsel = list(inprior = 0.5)),
                warning = function(w) conditionMessage(w))

writeLines(strwrap(bad, 72))
```

The package default for `coef$v_i` is 0.01, which is past the bound, so a BVS
model that does not set `coef$v_i` will say so. A prior precision of zero, the
flat limiting prior the package otherwise allows, is refused outright: there
would be nothing to draw an excluded coefficient from.

Note also the column of inclusion probabilities in the table above. Even over
the range where the chain mixes, the probability falls as the prior widens,
from 0.17 to 0.03. That is Bartlett's paradox: the Bayes factor between the two
states depends on how spread out the prior on the coefficient is, and a prior
made vaguer penalises the larger model. It is a property of the algorithm, not
of this implementation, and it is the reason to report a BVS inclusion
probability together with the prior that produced it.

## Which of the two

|  | `"ssvs"` | `"bvs"` |
|---|---|---|
| Excluded coefficient | held near zero by a tight prior | exactly zero |
| What the prior does | the two-point mixture in `varsel` | the ordinary normal in `coef` |
| Needs a proper `coef$v_i` | no | yes, and on the scale of the data |
| Mixing under correlated regressors | better | worse |
| Sensitivity to the prior width | mild | Bartlett's paradox |

SSVS is the safer choice when the prior was not picked with the selection in
mind, and the better one when the regressors are correlated, since nothing has
to travel between the prior and the data to change state. BVS gives exact
zeros, a frontier that genuinely omits a regressor, and an indicator that needs
no threshold to interpret.

Where the answer matters, run both. They agree here:

```{r}
rbind(ssvs = round(colMeans(as.matrix(ssvs$posterior$inclusion$coeffs)), 3),
      bvs = round(colMeans(as.matrix(bvs$posterior$inclusion$coeffs)), 3))
```

## Choosing the candidates

Both algorithms take the same three arguments for this. `include` names the
columns of the design matrix to select on, by name or by position;
`exclude_intercept` drops the intercept from whatever that leaves, which it
does by default; and `inprior` sets the prior inclusion probability, either one
number or one per candidate.

```{r}
picked <- create_sfmodel_exp(y ~ x1 + x2 + z1 + z2,
                             data = d,
                             varsel = "ssvs",
                             iterations = 1000,
                             burnin = 500)

picked <- add_priors(picked,
                     varsel = list(tau = c(0.01, 10),
                                   include = c("z1", "z2"),
                                   inprior = 0.2))

picked$priors$varsel$names
```

A regressor the frontier is known to need should be left out of the selection
rather than given a high `inprior`: the first states the knowledge, the second
only tilts the search.

## What the selection does not touch

Variable selection applies to the frontier. The inefficiency term, the error
and the augmented draws are sampled exactly as they are without it, so the
efficiency scores are read off the draws in the usual way, with the one
difference that they are now averaged over the frontiers the selection admits
rather than conditioned on one:

```{r}
plain <- create_sfmodel_exp(y ~ x1 + x2 + z1 + z2,
                            data = d,
                            iterations = 4000,
                            burnin = 1000)

plain <- add_posterior_coefficients(add_priors(add_seed(plain, 1234)))

round(c(none = mean(efficiency(plain)$mean),
        ssvs = mean(efficiency(ssvs)$mean),
        bvs = mean(efficiency(bvs)$mean)), 4)
```

The scores barely move, which is what to expect when the regressors being
dropped were noise: they were contributing nothing to the frontier, so removing
them takes nothing away from it. The gain is in the coefficients and in not
having had to decide by hand.

## The four-component model

`varsel` sits in the same place in `create_sfmodel4_exp()` and
`create_sfmodel4_hn()`, and means the same thing. The selection is on the
frontier, so the unit effect and the two one-sided terms are untouched by it.

```{r}
dp <- sim_sf4(n = 50, n_time = 8, beta = c(1, 0.5, 0.3))
dp$z1 <- rnorm(nrow(dp))

m4 <- create_sfmodel4_exp(y ~ x1 + x2 + z1, data = dp, id = "id",
                          varsel = "ssvs", iterations = 2000, burnin = 1000)
m4 <- add_priors(m4, varsel = list(semiautomatic = c(0.1, 10)))
m4 <- add_posterior_coefficients(add_seed(m4, 1234))

round(colMeans(as.matrix(m4$posterior$inclusion$coeffs)), 3)
```

## Information criteria

`selection_criteria()` works as it always does, with one caveat. AIC, BIC and
HQ charge the model for every coefficient the frontier was written with, not
for the number the selection kept, because that number is not the same in every
draw and there is no whole number to subtract. WAIC reads the effective number
of parameters off the draws and accounts for the selection by itself, so it is
the one to compare a selected model with an unselected one on.

```{r}
crit <- function(m) selection_criteria(add_posterior_loglik(m))
round(c(none = crit(plain)$WAIC$mean, ssvs = crit(ssvs)$WAIC$mean,
        bvs = crit(bvs)$WAIC$mean), 2)
```

## References

George, E. I., Sun, D., & Ni, S. (2008). Bayesian stochastic search for VAR
model restrictions. *Journal of Econometrics*, 142(1), 553–580.

Korobilis, D. (2013). VAR forecasting using Bayesian variable selection.
*Journal of Applied Econometrics*, 28(2), 204–230.

van den Broeck, J., Koop, G., Osiewalski, J., & Steel, M. F. J. (1994).
Stochastic frontier models: A Bayesian perspective. *Journal of Econometrics*,
61(2), 273–303.
