---
title: "Supervisory data for a stochastic frontier model of bank lending capacity"
subtitle: "Sources under the ITS on supervisory reporting, variable construction, and the specifications to estimate"
author: "Franz X. Mohr"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    embed-resources: true
---

# Purpose and scope

The question is whether banks are constrained in their new lending by regulatory
capital. The design is a stochastic frontier model in which the frontier is not
an estimated envelope of unknown economic content but the capital constraint
itself: given a bank's capital headroom and the risk weight density of the
lending it originates, there is an arithmetic maximum to how much new credit it
can extend. What the model estimates is the distance to that maximum, and how
that distance behaves as headroom shrinks.

The unit of observation is **bank × lending segment × quarter**. Segments are
kept apart — non-financial corporations, household mortgages, household
consumer credit — because their origination technologies and risk weights differ
enough that pooling them imposes a technology that does not exist.

This document covers three things: which supervisory sources supply each
variable, how each variable is computed from them, and which specifications to
estimate in what order.

# Resolving the reporting positions

The ITS on supervisory reporting is **Commission Implementing Regulation (EU)
2024/3117**, which replaced Regulation (EU) 2021/451. Reporting framework
release 4.0 applies from the March 2025 reference date and 4.2 from the fourth
quarter of 2025; xBRL-CSV becomes the mandatory transmission format from the
31 March 2026 reference date.

The tables below give, for every variable, the **template** and the **item as
it is labelled in that template**. That pair is what identifies a position, and
both are stable across releases.

Row and column *codes* are not stable — they shift with each amendment, and a
row number that was right for framework 3.2 can point at a different datapoint
in 4.2. They are therefore not asserted here. The one artefact that resolves
them all, for a given release, is:

> **`annotated_templates.zip`** — "DPM table layout and data point
> categorization", published on the EBA reporting framework page for each
> release. It maps template + row + column to datapoint identifiers.

The working method is: take the template and item label from the tables below,
look the pair up in the annotated table layout for the release in force, and
record the resulting row, column and datapoint ID in a version-controlled
mapping table. The extraction code then reads that table rather than carrying
literal coordinates, which is what makes a framework upgrade a data change
rather than a code change.

Supporting artefacts on the same page, in decreasing order of usefulness here:
the **DPM database** (4.0 in versions 1.0 and 2.0) for programmatic lookups,
the **DPM dictionary glossary** for the item definitions, the **validation
rules** packages, and `eba_identical_cells_4.x.xlsx`, which lists datapoints
reported in more than one place and is worth consulting before choosing between
two templates that appear to carry the same figure.

# Reporting frameworks in scope

| Framework | Legal act | Frequency | Reporting level | Used for |
|---|---|---|---|---|
| COREP Own Funds (C 01.00–C 05.01) | CIR (EU) 2024/3117 | Quarterly | Consolidated and solo | Own funds, requirements, ratios, buffers |
| COREP Credit Risk SA (C 07.00) | CIR (EU) 2024/3117 | Quarterly | Consolidated and solo | Exposure and RWA by exposure class, standardised |
| COREP Credit Risk IRB (C 08.01, C 08.02) | CIR (EU) 2024/3117 | Quarterly | Consolidated and solo | Exposure, RWA, PD/LGD bands, internal models |
| COREP Countercyclical buffer (C 09.02, C 09.04) | CIR (EU) 2024/3117 | Quarterly | Consolidated | Institution-specific countercyclical buffer rate |
| COREP Leverage (C 47.00) | CIR (EU) 2024/3117 | Quarterly | Consolidated and solo | Leverage exposure measure, the competing constraint |
| COREP LCR (C 72.00–C 76.00) | CIR (EU) 2024/3117 | Monthly | Consolidated and solo | Liquidity buffer, net outflows |
| COREP ALMM (C 66.00–C 71.00) | CIR (EU) 2024/3117 | Monthly/quarterly | Consolidated and solo | Funding concentration, roll-over, maturity ladder |
| COREP NSFR (C 80.00–C 84.00) | CIR (EU) 2024/3117 | Quarterly | Consolidated and solo | Available and required stable funding |
| COREP Large exposures (C 26.00–C 30.00) | CIR (EU) 2024/3117 | Quarterly | Consolidated | Concentration headroom |
| FINREP (F 01.xx, F 05.01, F 06.01, F 18.00, F 20.xx) | CIR (EU) 2024/3117, Annexes III–IV | Quarterly | Consolidated | Balance sheet, loans by product and sector, non-performing exposures, geography |
| AnaCredit | Regulation (EU) 2016/867 (ECB/2016/13) | Monthly | **Observed agent / legal entity** | Loan-level new business, borrower identity, PD, collateral |
| MIR | Regulation (EU) No 1072/2013 (ECB/2013/34) | Monthly | Solo, MFI | New business volumes and rates, including households |
| BSI | ECB balance-sheet-items regulation | Monthly | Solo, MFI | Stocks and reclassification-adjusted flows |
| SREP outcomes (P2R, P2G) | Supervisory decisions | Annual, ad hoc | Consolidated | Bank-specific capital requirements |
| Macroprudential measures | ESRB notifications, national acts | Event | Jurisdiction | CCyB, O-SII, SyRB, borrower-based measures |

Two of these are not reporting templates at all and are the ones most often
missing from a first extraction: **SREP outcomes**, which arrive as supervisory
decisions rather than as a reported datapoint, and **macroprudential measures**,
which have to be assembled from the ESRB notification database and national
legal acts and then mapped to the bank's exposure geography.

# The dependent variable: gross new lending

## Definition

Gross new business originated by bank $i$ in segment $s$ during quarter $t$:
the sum of committed amounts on credit agreements whose inception falls in the
quarter. **Gross, not net.** Net flows go negative and cannot be logged, and
more importantly a net flow confuses origination with amortisation of the
existing book, which is not a decision the bank takes in the quarter.

## From AnaCredit

AnaCredit is the preferred source because it identifies the borrower, which is
what makes the demand controls and the Khwaja–Mian cross-check possible.

$$
\text{NL}_{ist} = \sum_{j \,\in\, \mathcal{J}_{ist}} \text{CommitmentAmount}_{j}
$$

where $\mathcal{J}_{ist}$ is the set of instruments reported by bank $i$ with
*inception date* in quarter $t$, whose borrower falls in segment $s$.

Construction notes, each of which changes the number materially:

- **Commitment amount at inception**, from the instrument dataset, not the
  outstanding nominal amount, which is already amortised by the first reporting
  date and understates origination.
- **Exclude renegotiations.** A renegotiated instrument gets a new inception
  date in some reporting practices. Identify them through the instrument's
  status and the counterparty-instrument linkage and treat them as a separate
  output, since a renegotiation is not new credit.
- **Exclude revolving redraws.** Drawings under an existing credit line are not
  a new instrument, but a newly *granted* line is. Decide explicitly and record
  the decision: this is the single largest source of divergence between two
  analysts' series.
- **Exclude intra-group and purchased instruments**, which are not origination.
- **Mind the reporting threshold.** AnaCredit covers instruments with a
  commitment of €25,000 or more towards a given debtor, and covers **legal
  entities only** — households are out of scope entirely.
- **Segment from the counterparty reference data**: institutional sector and
  NACE code for corporates, and, for the household segments, AnaCredit cannot
  help at all.

## From MIR, for the household segments

MFI interest rate statistics report new business volumes by instrument category
including households, which is the only routine source for mortgage and consumer
credit origination. The cost is that there is no borrower identity, so
firm × time fixed effects are unavailable and demand has to be controlled for
with regional aggregates instead.

If both sources are used, the household and corporate segments rest on
different data and should never be pooled into a single frontier.

## Zeros and sparse activity

A bank with no origination in a segment in a quarter contributes a zero, which
the log cannot take. Either restrict the estimation sample to banks
continuously active in the segment and say so, or use a sample-selection
frontier. Dropping the zeros without comment biases the efficiency
distribution upward, because inactivity and low utilisation are the same
phenomenon at the margin.

# The capital-implied maximum

This is the construction that turns a generic frontier into a test of the
capital constraint.

## Capital headroom in euros

Headroom is the CET1 a bank holds in excess of what it must hold, measured at
the point where a shortfall has consequences — the Maximum Distributable Amount
trigger, which is a breach of the **combined buffer requirement**.

$$
H_{it} \;=\; \text{CET1}_{it} \;-\; \bigl(\text{P1R}^{\text{CET1}} + \text{P2R}^{\text{CET1}}_{it} + \text{CBR}_{it}\bigr)\times \text{RWA}_{it}
$$

with the combined buffer requirement

$$
\text{CBR}_{it} = \text{CCoB} + \text{CCyB}_{it} + \max\bigl(\text{O\text{-}SII}_{it},\, \text{G\text{-}SII}_{it},\, \text{SyRB}_{it}\bigr)
$$

subject to the stacking rules in force, which have changed and are not uniform
across jurisdictions.

| Component | Template | Item label to look up |
|---|---|---|
| CET1 capital | C 01.00 | *Common Equity Tier 1 capital* |
| Tier 1 capital | C 01.00 | *Tier 1 capital* |
| Total own funds | C 01.00 | *Own funds* |
| Risk exposure amount | C 02.00 | *Total risk exposure amount* |
| Capital ratios | C 03.00 | *CET1 capital ratio*; *Tier 1 capital ratio*; *Total capital ratio* |
| Surplus over requirement | C 03.00 | *Surplus(+)/Deficit(-) of CET1 capital* |
| Combined buffer requirement | C 04.00 | *Combined buffer requirement* |
| Buffer components | C 04.00 | *Capital conservation buffer*; *Institution specific countercyclical capital buffer*; *Systemic risk buffer*; *Global/Other Systemically Important Institution buffer* |
| Capital left over after buffers | C 04.00 | *CET1 capital available to meet buffers* |
| Pillar 2 add-on as reported | C 04.00 | *Own funds requirements related to Pillar II adjustments* |
| Institution-specific CCyB rate | C 09.04 | *Institution specific countercyclical buffer rate* |
| Geographical distribution behind that rate | C 09.02 | *Own funds requirements*, by country of the exposure |
| Pillar 2 requirement, CET1 portion and composition | Not a reported position | SREP decision; disclosed under Art. 447 CRR |

*CET1 capital available to meet buffers* in C 04.00 deserves particular
attention: it is the supervisor's own answer to the question the headroom
formula asks, already net of the amounts absorbed by the Pillar 1 and Pillar 2
requirements. Computing headroom independently and reconciling it against this
item is the cheapest validation available.

Three traps:

1. **Use the CET1 portion of P2R.** Since CRD V, P2R is met with a prescribed
   composition (at least 56.25% CET1). Applying the whole P2R against CET1
   overstates the constraint.
2. **AT1 and T2 shortfalls migrate upward.** If a bank does not fill its AT1 or
   T2 buckets, the gap must be met with CET1, so the effective CET1 requirement
   exceeds the nominal one. This is precisely the bank that is constrained, so
   ignoring it removes the signal.
3. **P2G is not part of the MDA stack** but banks manage to it. Compute headroom
   both ways — to the MDA trigger and to P2R + CBR + P2G — and treat the
   difference as a specification check rather than choosing one.

## Risk weight density of new lending

Capital does not constrain volume; it constrains risk-weighted assets. The
bridge between the two is the risk weight density of *newly originated* lending
in the segment, which is bank-specific and is not the bank's average density.

For IRB banks, derive it from the loan-level risk parameters of the new
instruments:

$$
rw_{ist} \;=\; \frac{\sum_{j \in \mathcal{J}_{ist}} w\bigl(\text{PD}_j, \text{LGD}_j, M_j, S_j\bigr)\cdot \text{EAD}_j}{\sum_{j \in \mathcal{J}_{ist}} \text{EAD}_j}
$$

with $w(\cdot)$ the CRR supervisory formula for the relevant exposure class.
For standardised banks the weight follows from the exposure class and, where
applicable, the loan-to-value band. COREP C 08.01 (IRB) and C 07.00 (SA) give
the aggregate cross-check: the density implied by the new-business calculation
should sit in the neighbourhood of the reported density for the same exposure
class.

::: {.callout-warning}
## IRB and SA banks are not comparable

The same exposure attracts a materially different risk weight under the two
approaches, so the capital-implied maximum is not on a common scale across
banks with different approaches. Either estimate separate frontiers, or include
the approach as a frontier control and never interpret cross-approach
differences in the efficiency score. The output floor phase-in narrows this over
time, which is itself usable variation.
:::

## Putting them together

Capital consumed per euro of new lending, and the implied maximum:

$$
c_{ist} = rw_{ist}\times k_{it}, \qquad
\text{NL}^{\max}_{ist} = \frac{H_{i,t-1}}{c_{is,t-1}}
$$

where $k_{it}$ is the applicable total requirement ratio. A worked example, for
a bank with €800m of headroom lending to corporates at a 75% density under a
15% total requirement: $c = 0.75 \times 0.15 = 0.1125$, so
$\text{NL}^{\max} = 800/0.1125 \approx €7.1$bn of new corporate lending before
the buffer is exhausted.

# The remaining variables

| Variable | Template | Item label to look up | Construction |
|---|---|---|---|
| Stable funding | C 84.00 | *Available stable funding*; *Required stable funding*; *Net stable funding ratio* | Headroom as ASF − RSF, in euros |
| Customer deposits | F 08.01 | *Deposits*, by counterparty sector | Level, in logs |
| Liquidity headroom | C 76.00 | *Liquidity buffer*; *Net liquidity outflow*; *Liquidity coverage ratio* | $\text{LB} - \text{NLO}$, in euros, not the ratio |
| Leverage headroom | C 47.00 | *Tier 1 capital*; *Total exposure measure*; *Leverage ratio* | $\text{T1} - r\times\text{LEM}$ at the applicable requirement |
| Concentration headroom | C 26.00–C 30.00 | *Exposure value*, by counterparty, against the limit | Binding for lending to existing large borrowers |
| Loan stock, by product | F 05.01 | *Loans and advances*, by product | Level, in logs, lagged |
| Loan stock, corporates by sector | F 06.01 | *Loans and advances to non-financial corporations*, by NACE | Segment-level scale |
| Risk weights, standardised | C 07.00 | *Exposure value*; *Risk weighted exposure amount*, by exposure class | Density by class, cross-check on $rw$ |
| Risk weights, internal models | C 08.01 | *Exposure value*; *Risk weighted exposure amount*; *Exposure weighted average PD*; *Exposure weighted average LGD* | Density and risk parameters by class |
| Operating and staff expenses | F 02.00 | *Administrative expenses*; of which *Staff expenses* | The labour input; FTE is **not** an ITS position |
| Credit quality of the book | F 18.00 | *Performing exposures*; *Non-performing exposures*; *Accumulated impairment* | NPL ratio and coverage |
| Risk of new lending | AnaCredit | Instrument and counterparty risk datasets | Exposure-weighted PD over $\mathcal{J}_{ist}$ |
| Demand | AnaCredit + national accounts | Counterparty reference data for the geography | $\sum_r \omega_{irt}\,\Delta\log Y_{rt}$, $\omega$ the lagged exposure share |
| Policy rate, term spread | Not reported | ECB and market data | Common across banks, absorbed by time effects |
| Distance to MDA | Derived | See the headroom table | The key inefficiency determinant |
| P2R, CCyB, borrower-based measures | Not reported as such | SREP decisions, ESRB notifications, national acts | Level and change; the identifying variation |

Three remarks on this table.

**Four of these are not reporting positions at all.** Full-time equivalents,
the policy rate, the Pillar 2 requirement and the macroprudential measures have
no datapoint in the ITS. FTE comes from national returns or annual accounts,
which is a good reason to use *Administrative expenses* from F 02.00 as the
labour input instead; P2R arrives as a supervisory decision and is disclosed
under Article 447 CRR; the macroprudential measures have to be assembled from
ESRB notifications and national legal acts and then mapped to the bank's
exposure geography. Budget for these separately — they are the positions a
first extraction always misses, and two of them carry the identifying
variation.

**The demand control is the single most important variable in the design.**
Without it, a quarter in which nobody wanted to borrow is scored as bank
inefficiency and the model measures nothing.

**Check `eba_identical_cells` before choosing a template.** Several of these
figures are reported in more than one place — own funds items appear in both
C 01.00 and the ratios template, loan stocks in both F 01.01 and F 05.01 — and
the identical-cells file tells you which are genuinely the same datapoint and
which only look alike.

# Alignment: what breaks when the sources are joined

::: {.callout-important}
## The consolidation mismatch is the serious one

COREP and FINREP are reported on a **consolidated** basis by the group.
AnaCredit is reported by **observed agents**, which are legal entities —
typically the institution and its foreign branches, not the consolidated group.
A capital headroom measured for the group and a new-lending volume measured for
one legal entity within it are not the same bank.

There is no clean fix. The options are: restrict to banks where the two
perimeters effectively coincide, aggregate AnaCredit observed agents up to the
group using the group structure in FINREP F 40.xx, or run the analysis at solo
level using solo COREP where it is reported. Whichever is chosen has to be
stated, because the choice moves the results.
:::

Further alignment work, all of it unglamorous and all of it necessary:

- **Frequency.** AnaCredit and MIR are monthly, COREP and FINREP quarterly.
  Aggregate the flows to quarters rather than interpolating the stocks.
- **Reference date versus remittance date.** Capital variables must be lagged
  to the previous quarter-end anyway for predeterminedness, which conveniently
  also resolves the fact that a bank does not know its own quarter-end capital
  when it originates during the quarter.
- **Restatements.** Supervisory data is resubmitted. Vintage the extraction and
  record the pull date, or results will not reproduce.
- **Mergers and entity changes.** An LEI change or a merger breaks the panel.
  Build an entity-mapping table and decide whether the merged entity is a new
  unit or a continuation; for the four-component model this matters a great
  deal, since a spurious break shows up as persistent inefficiency.
- **Accounting framework.** FINREP has separate annexes for IFRS and national
  GAAP reporters. Some items are not comparable across them.
- **Currency.** Convert at the ECB reference rate for the reference date;
  do not mix translation methods between the numerator and denominator of a
  ratio.

# How the variables enter the model

Two forms are worth estimating, and they differ in whether the capital-implied
maximum is imposed or estimated.

## Estimated frontier

$$
\ln \text{NL}_{ist} = f(x_{i,t-1};\beta) + d_{ist}'\gamma + \tau_t + \mu_i - \eta_i + v_{ist} - u_{ist}
$$

with $x$ the capacity inputs (capital headroom, stable funding, liquidity
headroom, labour, loan stock) in logs, and $d$ the demand and risk controls.

## Imposed frontier: capacity utilisation

$$
\ln\!\left(\frac{\text{NL}_{ist}}{\text{NL}^{\max}_{is,t-1}}\right) = d_{ist}'\gamma + w_{ist}'\theta + \tau_t + \mu_i - \eta_i + v_{ist} - u_{ist}
$$

Here the dependent variable is already a utilisation rate, $w$ holds the
*competing* constraints (funding, liquidity, leverage, concentration), and
$\exp(-u)$ is capital-capacity utilisation directly.

## Placement rules

| Where | What goes there | Why |
|---|---|---|
| Frontier $f(\cdot)$ | Capacity inputs, lagged | These define the maximum |
| Frontier controls $d$ | Demand, policy rate, competition, risk of new lending | Otherwise weak demand is scored as inefficiency |
| Competing constraints $w$ | Funding, liquidity, leverage, concentration headroom | So that capital is not the only candidate the model can find |
| Inefficiency mean | Distance to MDA, P2R, CCyB, borrower-based measures, NPL ratio, TLTRO participation | The policy question |
| $\sigma_u$ scale function | Distance to MDA, log assets | Where the binding-constraint signature appears |
| $\sigma_v$ scale function | Log assets | Bank size spans orders of magnitude |

Everything on the right-hand side that the bank chooses enters **lagged one
quarter**. Capital headroom in particular is chosen in anticipation of lending,
so contemporaneous headroom is endogenous by construction.

**Estimate on log output.** The efficiency score $\exp(-u)$ is a proportional
shortfall only when the dependent variable is a log; fitting euro volumes makes
the results depend on whether the data are in euros or millions.

# Specifications to estimate

The ladder below runs from diagnostic to answer. Each step is informative about
whether the next one is worth running.

## S0. Pooled cross-section, diagnostic only

```{r}
#| eval: false
library(bsfa)

s0 <- create_sfmodel_exp(log_nl ~ log_caphead + log_deposits + log_opex +
                           log_stock + demand + avg_pd,
                         data = panel, iterations = 20000, burnin = 2000)
s0 <- add_posterior_loglik(add_posterior_coefficients(add_priors(s0)))
summary(s0)
selection_criteria(s0)
```

Purpose: confirm the composed error is identified at all. If the posterior for
`lambda` piles up against zero, the one-sided term is not separating from noise
at this frequency — aggregate to semi-annual or annual before going further.
This is the only step at which the information criteria are available, which is
also where the exponential and half-normal specifications should be compared.

## S1. Capacity-normalised utilisation, by quarter

The same call with the dependent variable replaced by
`log(nl / nl_max_lag)` and the capacity inputs replaced by the competing
constraints. Run it separately by quarter to trace utilisation through the
cycle without imposing a panel structure.

## S2. Time-invariant panel, structural ranking

```{r}
#| eval: false
s2 <- create_sfmodel_exp(log_nl ~ ..., data = panel, id = "bank_id",
                         iterations = 20000, burnin = 2000)
s2 <- add_posterior_coefficients(add_priors(s2, lambda = list(r_star = 0.85)))
efficiency(s2)
```

Purpose: a ranking of banks by structural origination capacity. **Not** an
answer to the capital question, because a constraint that binds cyclically has
nowhere to live in a model whose inefficiency term is constant per bank, and
because business-model heterogeneity is booked as inefficiency here.

## S3. Four-component — the main specification

```{r}
#| eval: false
s3 <- create_sfmodel4_exp(log_nl ~ ..., data = panel, id = "bank_id",
                          iterations = 40000, burnin = 5000)
s3 <- add_priors(s3,
                 lambda_eta = list(r_star = 0.95),   # persistent
                 lambda_u   = list(r_star = 0.90))   # transient
s3 <- add_posterior_coefficients(add_seed(add_initial_values(s3), 1))

efficiency(s3, type = "persistent")   # structural capacity, per bank
efficiency(s3, type = "transient")    # cyclical restraint, per bank-quarter
```

This is the specification the question needs. The mapping is:

- $\mu_i$ — business model, market, client base. Not inefficiency.
- $\eta_i$ — structural lending capacity.
- $u_{it}$ — **cyclical restraint, which is where a binding capital constraint
  appears.**

Then relate $\hat u_{it}$ to distance-to-MDA. Note the identification caveat:
what the data determine sharply is $\mu_i - \eta_i$, and the split between them
depends on which has the wider spread. Run the prior sensitivity check on
`sigma_mu` and on the persistent inefficiency parameter, and read the persistent
scores with that in mind. The transient scores, which carry the answer, are not
affected.

## S4. Heteroskedastic inefficiency — the actual test

The signature of a binding constraint is distributional: the dispersion of
inefficiency should collapse as headroom shrinks, because a bank pressed against
its limit has no room to vary.

$$
\sigma_{u,ist} = \exp\bigl(\psi_0 + \psi_1\,\text{MDAdist}_{i,t-1} + \psi_2 \ln \text{Assets}\bigr)
$$

A positive $\psi_1$ is the finding. A $\psi_1$ of zero says the frontier is
measuring something other than the capital constraint, whatever else it is
measuring. **Not yet implemented in `bsfa`** — see the roadmap.

## S5. Latent class — constrained and unconstrained regimes

Two classes with membership driven by distance to MDA, AT1 trigger proximity and
recent P2R changes, giving a posterior probability that each bank-quarter is
constrained and a direct estimate of the lending forgone. Capital constraints
are not linear — they do nothing until they bite — so a single linear
coefficient averages the effect towards zero. **Not yet implemented in `bsfa`.**

## S6. Within-firm cross-check

Not a frontier model. On the AnaCredit loan-level data, regress the change in
credit to firm $f$ from bank $i$ on bank capital with **firm × time fixed
effects**, which absorb credit demand entirely and identify supply from
variation across lenders to the same borrower.

This cannot give a level — firm × time effects difference out the aggregate —
but it gives a credible elasticity. Use it to validate S3 and S4: if the
frontier says capacity is slack while the within-firm estimator finds a strong
capital effect, the frontier is measuring something other than the capital
constraint, and $\psi_1$ from S4 will usually say why.

## Robustness

| Check | What it tests |
|---|---|
| Exponential versus half-normal | Sensitivity of the scores to the one-sided distribution |
| `r_star` at 0.75, 0.85, 0.95 | Prior sensitivity of the efficiency level |
| Gamma prior on `sigma_mu` varied | Whether the S3 split is data or prior |
| Headroom to MDA versus to P2G | Whether banks manage to the regulatory or the market requirement |
| IRB and SA banks separately | Whether the density construction is comparable |
| Semi-annual aggregation | Whether the quarterly signal-to-noise ratio is adequate |
| Excluding renegotiations versus including | The largest construction choice in the dependent variable |

# Pre-estimation checks

- Reconcile aggregated AnaCredit new business against MIR new business volumes
  for the overlapping segment. A persistent gap means the exclusions are wrong.
- Reconcile derived capital headroom against reported capital ratios in
  C 03.00; they must imply each other.
- Check that the new-lending risk weight density sits near the reported density
  for the same exposure class.
- Plot $\text{NL}/\text{NL}^{\max}$ before estimating anything. Values above one
  mean the headroom or the density is wrong, not that the bank has broken the
  rules.
- Count banks by quarter and confirm the panel is not silently unbalanced by
  entity changes.
- Confirm that the consolidation perimeter used is the same on both sides of
  every ratio.

# What the package supports today

| Specification | Status |
|---|---|
| S0, S1 cross-section | `create_sfmodel_exp()`, `create_sfmodel_hn()` |
| S2 time-invariant panel | Same, with `id` |
| S3 four-component | `create_sfmodel4_exp()`, `create_sfmodel4_hn()` |
| S4 heteroskedastic $\sigma_u$ | Not implemented |
| S5 latent class | Not implemented |
| Information criteria | Cross-section only; unavailable for panel and four-component models |

# References

Aigner, D., Lovell, C. A. K., & Schmidt, P. (1977). Formulation and estimation
of stochastic frontier production function models. *Journal of Econometrics*,
6(1), 21–37.

Battese, G. E., & Coelli, T. J. (1995). A model for technical inefficiency
effects in a stochastic frontier production function for panel data.
*Empirical Economics*, 20(2), 325–332.

Commission Implementing Regulation (EU) 2024/3117 laying down implementing
technical standards for the application of Regulation (EU) No 575/2013 with
regard to supervisory reporting of institutions.

Khwaja, A. I., & Mian, A. (2008). Tracing the impact of bank liquidity shocks:
Evidence from an emerging market. *American Economic Review*, 98(4), 1413–1442.

Kumbhakar, S. C., Lien, G., & Hjalmarsson, L. (2014). Technical efficiency in
competing panel data models. *Journal of Productivity Analysis*, 41(2),
321–337.

Regulation (EU) 2016/867 of the European Central Bank on the collection of
granular credit and credit risk data (ECB/2016/13).

Regulation (EU) No 1072/2013 of the European Central Bank concerning statistics
on interest rates applied by monetary financial institutions (ECB/2013/34).

van den Broeck, J., Koop, G., Osiewalski, J., & Steel, M. F. J. (1994).
Stochastic frontier models: A Bayesian perspective. *Journal of Econometrics*,
61(2), 273–303.
